---
title: "LakeMichigan_ASV_Biodiversity_Analysis"
author: "Daniyal Tariq"
date: '`r Sys.Date()`'
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Set Seed

```{r set-seed}
set.seed(238428)
```

##Load Libraries

```{r load-libraries}
pacman::p_load(tidyverse, devtools, patchwork, iNEXT, phyloseq,
               install = FALSE)
```

## Load data
```{r load-data}

load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

```{r}
min(sample_sums(raw_preprocessed_physeq))

metadata_df <-
  raw_preprocessed_physeq %>%
  sample_data() %>%
  data.frame()

head(metadata_df)
```

```{r setting-lakesite-colors}
lakesite_colors <- c("110" = "dodgerblue4","M15" = "dodgerblue2","M45" = "#D9CC3C","MLB" = "#A0E0BA")
```


## Goals

1. Calculate the Hill Diversity of the samples.
2. Evaluate the rarefaction curves.
3. Evaluate the Diversity values.
4. Makes notes of specific samples and their seq depth.

##Diversity Calculations with iNEXT

```{r making-iNEXT-df}
iNEXT_input_df <- 
  raw_preprocessed_physeq %>%
  otu_table() %>%
  data.frame()
# Quick check
dim(iNEXT_input_df)
```

## Running iNEXT for generating diversity numbers
```{r running-iNEXT}
#iNEXT_data <- iNEXT(iNEXT_input_df,q = c(0,1,2), datatype = "abundance")

# Save the file
#save(iNEXT_data, file = "data/04_Biodiversity/iNEXT_data.RData")
```


##Load data
```{r load-iNEXT-obj}
load("data/04_Biodiversity/iNEXT_data.RData")
str(iNEXT_data)
```

## Plot Diversity
```{r}
# Prepare Colors 
color_df <- 
  iNEXT_input_df %>%
  colnames() %>%
  data.frame()
# Check
head(color_df)
```


```{r}
# Rename the column 
colnames(color_df)[1] <- "names"
# Check
head(color_df)
```

```{r}
# Make a helper dataframe for plotting with colors 
iNEXT_color_df <- 
  color_df %>%
  # Fix the names for merging
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X",  replace = "")) %>%
  # Merge with metadata
  left_join(metadata_df, by = "names") %>%
  # Merge with colors for plotting with ggiNEXT
  left_join(data.frame(lakesite_colors = lakesite_colors,
            lakesite = names(lakesite_colors)),
            by = "lakesite")
```


## Plotting rarefaction with ggiNEXT
```{r plotting-rarefaction-curves}
ggiNEXT(iNEXT_data, type = 1, facet.var = "Order.q") + 
  facet_wrap(~Order.q, scales = "fixed") + 
  scale_color_manual(values = iNEXT_color_df$lakesite_colors, guide = FALSE) + 
  scale_fill_manual(values = iNEXT_color_df$lakesite_colors, guide = FALSE) + 
  scale_shape_manual(values = base::rep(17, nsamples(raw_preprocessed_physeq)),
                     guide = FALSE) +
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  theme(legend.position = "none") 
```

The rarefaction curves look very similar to those we generated in class. We note the following:

1.For each q value, we see plateaus for all the samples, indicating that the library size doesn't continue to influence the measures of diversity and we have an appropriate sequencing depth for our analysis. 
2.The curves follow the expected pattern with the richness measure at q =0, having the most ASVs. We also, see that there is not a high amount of variations between different samples from the same depth they were collected from. (Depth refers to the depth of the lake where the sample came from in this case)

## Manually plot Diversity

#Rarefaction

```{r}
iNEXT_manual_df <- 
  iNEXT_data$iNextEst$size_based %>%
  dplyr::rename(names = Assemblage) %>%
  # Fix the samples names 
  mutate(names = gsub(names, pattern = "[.]", replace = "-"),
         names = gsub(names, pattern = "X", replace = "")) %>%
  # join with metadata 
  left_join(., metadata_df, by = "names") %>%
  # Add colors to data frame
  left_join(., data.frame(lakesite_colors = lakesite_colors,
                          lakesite = names(lakesite_colors)),
            by = "lakesite") 
# Inspect 
dim(iNEXT_manual_df)
```


```{r}
# Plot it - Rarefaction Curve 
iNEXT_manual_df %>%
  # Filter out rows that are calcaulted by rarefaction from iNEXT
  dplyr::filter(Method == "Rarefaction") %>%
  # Now, let's draw the plot, be sure to group by the samples if using geom_line!
  ggplot(aes(x = m, y= qD, color = lakesite, group = names)) + 
  geom_line() + 
  # Facet with the station to see the samples better 
  facet_grid(Order.q~lakesite, scales = "free") + 
  scale_color_manual(values = lakesite_colors) + 
  labs(x = "Number of Sequences (Library Size)", 
       y = "Effective Number of ASVs") + 
  theme_bw() + 
  #guides(color = guide_legend(nrow = 2, byrow = TRUE)) + 
  theme(legend.position = "bottom",
        axis.text.x = element_text(angle = 30, vjust = 1, hjust = 1),
        legend.title = element_blank()) 
```


## Boxplots of Diversity
```{r}
# Prettier station labels 

lakesite_names <- c("110","M15","M45","MLB")

names(lakesite_names) <- c("110","M15","M45","MLB")

# Make a dataframe
obs_div_df <- 
  iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  left_join(data.frame(lakesite_names = lakesite_names, 
                       lakesite = names(lakesite_names)), 
                       by = "lakesite")

# Check it 
head(obs_div_df)
```

## 
```{r}
obs_div_df %>%
  ggplot(aes(x =lakesite_names, y = qD, fill = lakesite, color = lakesite)) + 
  facet_wrap(~Order.q, scales = "free") + 
  geom_jitter(size = 2.5) + 
  geom_boxplot(alpha = 0.5, outlier.shape = NA) + 
  scale_color_manual(values = lakesite_colors) + 
  scale_fill_manual(values = lakesite_colors) + 
  labs(y = "Effective Number of ASVs") + 
  theme_bw() +
  theme(legend.position = "bottom",
        axis.title.x = element_blank(),
        legend.title = element_blank())
```

## Looking at the effect of environmental variables
```{r}
head(metadata_df)

env_df <- 
  metadata_df %>%
  dplyr::select(names, limnion,season)
# inspect
head(env_df)
```

## Diversity as a function of season and lake depth
```{r}
iNEXT_manual_df %>%
  dplyr::filter(Method == "Observed") %>%
  ggplot(aes(x = season, y = qD)) + 
  facet_wrap(.~Order.q + lakesite, scales = "free") + 
  geom_boxplot(aes(color = lakesite))  + 
  stat_smooth(method = "lm", formula = y ~poly(x, 2)) + 
  labs(x = "Depth", y = "# of ASVs") + 
  scale_color_manual(values = lakesite_colors) + 
  theme_bw() + 
  theme(legend.position = "bottom",
        legend.title = element_blank())

```

We looked at each measure of diversity for different depths at different seasons. We note the following:

1.We see that each measure shows that at the deepest (the 110 sample) depth that was sampled, the diversity measures don't change that much and as we get closer to the surface, we see see a much greater seasonal dependence. This seems reasonable as the temperature of the water at large depths would not vary drastically between seasons and so the growth conditions for the microbes stay consistent.
2.Generally, we also see there is a much greater variance in the summer season at all depths, judging by the presence of many outliers and large inter-quartile ranges of the boxplots. 

