---
title: "Phylogenetic Tree Inspection and Rooting"
author: "Daniyal Tariq"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Goals of this file

1.Load the fastree unrooted tree.
2.Add tree to phyloseq object.
3.Visualize and inspect tree with ggtree.
4.Prune ASVs, if needed.
5.Root our tree.
6.Combine new tree with a phyloseq object.
7.Save 2 phyloseq objects: 
  1. Unrooted tree phyloseq object 
  2. Rooted tree phyloseq object.
  
##Set the seed

```{r}
set.seed(2443)
```

##Load Packages

```{r}
pacman::p_load(tidyverse, phyloseq, ggtree, phytools,
               install = FALSE)
```

##Load data files

```{r load-preprocessed-physeq}
# Preprocessed phyloseq object 
load("data/02_PreProcessing/raw_preprocessed_physeq.RData")
raw_preprocessed_physeq
```

```{r load-tree}

# Load in the tree! 
unrooted_tree <- read.tree("data/03_Phylogenetic_Tree/ASVs_unrooted.tree")
unrooted_tree
str(unrooted_tree)
```

##Merge Phyloseq with tree

```{r}
# Intuition check 
stopifnot(ntaxa(raw_preprocessed_physeq) == ntaxa(unrooted_tree))

# Merge the tree with the phyloseq object 
unrooted_physeq <- merge_phyloseq(raw_preprocessed_physeq, unrooted_tree)

unrooted_physeq
```


##Plot tree with `ggtree`

```{r plot-tree}
# Make a basic tree
kingdom_tree <- 
  ggtree(unrooted_physeq) + 
  # color tips by kingdom 
  geom_tippoint(mapping = aes(color = Kingdom)) + 
  scale_color_manual(values = c("goldenrod1", "cornflowerblue", "grey")) +
  # Add title 
  labs(title = "Unrooted Tree") + 
  #move the legend to the bottom 
  theme(legend.position = "bottom"); kingdom_tree

kingdom_tree
```

```{r add-node-labels-to-tree}
kingdom_node_tree <- 
  kingdom_tree + 
  # Add the node label 
  geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)
kingdom_node_tree
```

```{r evaluating-node-1721}
# View a specific clade 
# Zoom in on origin tree: Node 1721
viewClade(kingdom_node_tree + 
          labs(title = "Unrooted Tree: Node 1721"), 
          node = 1721)
```

```{r}

# View a specific clade 
# Zoom in on origin tree: Node 1743
viewClade(kingdom_node_tree + 
          labs(title = "Unrooted Tree: Node 1743"), 
          node = 1743)

```

```{r}
viewClade(kingdom_node_tree + 
          labs(title = "Unrooted Tree: Node 4605") + 
          geom_text(aes(label=ASV)), 
          node = 4605)
```

```{r}
unrooted_physeq %>%
  subset_taxa(., ASV == "ASV_1167") %>%
  tax_table() %>%
  data.frame()
```

We looked at both ASV 375 and 1167 because of the length of the node, but after blasting the sequence, it appears they just originate from the 16S gene of uncharacterized bacteria so we will not be removing these sequences.

##Midroot Tree

```{r}
new_unrooted_tree <- phy_tree(unrooted_physeq)
is.rooted(new_unrooted_tree)
```

```{r}
# Let's midpoint root the tree
midpoint_rooted_tree <- midpoint.root(new_unrooted_tree)

# Is the new tree rooted?
is.rooted(midpoint_rooted_tree)
```

```{r}

# Merge tree with the physeq
midroot_physeq <- merge_phyloseq(raw_preprocessed_physeq, midpoint_rooted_tree)
```

```{r}
midroot_physeq

# Quick inspection of tree 
ggtree(midroot_physeq) + 
  geom_tippoint(mapping = aes(color = Kingdom)) + geom_text(aes(label=node), hjust= -0.5, vjust = -0.3, size = 2)

```

##Save to a new phyloseq object

```{r}
# Save both phyloseq objects with our tree object to one .RData file 
save(list = c("unrooted_physeq", "midroot_physeq"),
     file = "data/03_Phylogenetic_Tree/phytree_preprocessed_physeq.RData")
```

##Session Info

```{r session-information}

# Ensure reproducibility 
devtools::session_info()

```

